Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-23 LICS_2025/monoidal_e_graphs/sections/appendix.tex
last section left for closed

    Then, faithfulness amounts to showing that if $\llbracket f \rrbracket \Rrightarrow_{\llbracket \mathcal{E} \rrbracket}^{*} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$ then $f = g$ modulo $\mathcal{E}$ in $\catname{S}^{+}(\Sigma)$.
% By the theorem~\ref{thm:completeness_simple} if $\llbracket f \rrbracket = \llbracket g \rrbracket$ via rules from $\mathcal{S}$ then $f = g$, so we will show that the same holds when $\llbracket f \rrbracket = \llbracket g \rrbracket$ via rules from $\mathcal{E}$.
Consider a rewrite rule $\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle = \langle \mathcal{L}, \mathcal{R} \rangle$ such that $\llbracket f \rrbracket \Rrightarrow_{\langle \mathcal{L}, \mathcal{R} \rangle} \llbracket g \rrbracket$ and suppose
\[
	\llbracket f \rrbracket = n \xrightarrow{} n' \xrightarrow{} \mathcal{F} \xleftarrow{} m' \xleftarrow{} m ~.
\]
This implies the existence of a diagram as in~\ref{def:dpoi-e}.
We will proceed by induction on the level of the occurrence of $\mathcal{L}$ within $\mathcal{F}$.
\begin{itemize}
	\item Suppose the image of $\mathcal{L}$ in $\mathcal{F}$ is top-level.
		  Then, by lemma~\ref{lemma:decomposition}, there exists a decomposition of $\llbracket f \rrbracket$ in terms of MDA cospans
		\begin{align*}
			(k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
			n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
			(i \xrightarrow{} i' \xrightarrow{} &\mathcal{L} \xleftarrow{} j' \xleftarrow{} j)
		\end{align*}
		By fullness of $\llbracket - \rrbracket$ (from Theorem~\ref{thm:completeness_simple}) there exist $c_{1}$ and $c_{2}$ such that 
		\begin{align*}
		&\llbracket c_{1} \rrbracket = n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k\\
        &\llbracket c_{2} \rrbracket = j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m
        \end{align*}
		By functoriality, $\llbracket f \rrbracket = \llbracket c_{1};(id \otimes l);c_{2} \rrbracket$ and by faithfulness we have that $f = c_{1};(id \otimes l);c_{2}$ to which we can apply $\langle l, r \rangle$ and get $g' = c_{1};(id \otimes r);c_{2}$.
		We then have $\llbracket g' \rrbracket = \llbracket c_{1};(id \otimes r);c_{2} \rrbracket =  \llbracket g \rrbracket$ since the boundary complement is unique (Proposition ~\ref{prop:boundary_unique}) and $g' = g$ by faithfulness.

		\item If the image of $\mathcal{L}$ is not top-level, then either $\llbracket f \rrbracket$ is decomposable as
		\begin{align*}
			(k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
			n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
			(i \xrightarrow{} i' \xrightarrow{} &\mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)
		\end{align*}
		such that $\mathcal{L}$ is a convex down-closed sub-e-hypergraph of $F'$.
		Since the image of $\mathcal{L}$ lies within $\mathcal{F}'$, there is an EDPO rewrite step that turns
		\[
			(i \xrightarrow{} i' \xrightarrow{} \mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)	
		\]
		into
		\[
			(i \xrightarrow{} i'' \xrightarrow{} \mathcal{G'} \xleftarrow{} j'' \xleftarrow{} j)	
		\]
		via $\langle \mathcal{L}, \mathcal{R} \rangle$.
		By fullness and inductive hypothesis, there exist $f'$ and $g'$ such that 
		\begin{align*}
			&\llbracket f' \rrbracket = i \xrightarrow{} i' \xrightarrow{} \mathcal{F'} \xleftarrow{} j' \xleftarrow{} j\\
            &\llbracket g' \rrbracket = (i \xrightarrow{} i'' \xrightarrow{} \mathcal{G'} \xleftarrow{} j'' \xleftarrow{} j)
        \end{align*}
		and that $f' = g'$.
		By decomposition and fullness we have $\llbracket f \rrbracket = \llbracket c_{1};(id \otimes f');c_{2} \rrbracket$ and $f = c_{1};(id \otimes f');c_{2}$ by faithfulness.
		By applying $f' = g'$ we get $f = g'' = c_{1};(id \otimes g');c_{2}$.
		Finally, $\llbracket g'' \rrbracket = \llbracket c_{1};(id \otimes g');c_{2} \rrbracket = \llbracket g \rrbracket$ (by the uniqueness of the boundary complement) and $g'' = g = f$ by faithfulness.

		\item or, $\llbracket f \rrbracket$ is decomposable as
		\begin{align*}
			n_1 \xrightarrow{} n_1' \xrightarrow{} &\mathcal{F}_1 \xleftarrow{} m_1' \xleftarrow{} m_1\\
			&\;+\\
			&\vdotswithin{+}\\
			&\;+\\
			n_p \xrightarrow{} n_p' \xrightarrow{} &\mathcal{F}_p \xleftarrow{} m_p' \xleftarrow{} m_p\\
			&\;+\\
			&\vdotswithin{+}\\
			&\;+\\
			n_k \xrightarrow{} n_k' \xrightarrow{} &\mathcal{F}_k \xleftarrow{} m_k' \xleftarrow{} m_k
		\end{align*}
		such that the image of $\mathcal{L}$ lies within $\mathcal{F}_{p}$.
		By inductive hypothesis we have that $\llbracket f_{p} \rrbracket = n_p \xrightarrow{} n_{p}' \xrightarrow{} \mathcal{F}_{p} \xleftarrow{} m_{p}' \xleftarrow{} m_{p}$ rewrites into $\llbracket g_{p} \rrbracket = n_p \xrightarrow{} n_{p}'' \xrightarrow{} \mathcal{G}_{p} \xleftarrow{} m_{p}'' \xleftarrow{} m_{p}$ such that $f_{p} = g_{p}$.
		By fullness and the fact that $\llbracket - \rrbracket$ is enriched, there exist $\llbracket f_1 + \ldots + f_{p} + \ldots + f_{k} \rrbracket = \llbracket f_{1} \rrbracket + \ldots + \llbracket f_{p} \rrbracket + \ldots + \llbracket f_{k} \rrbracket = \llbracket f \rrbracket$.
		By applying $f_{p} = g_{p}$ we get $f = g''$ and $\llbracket g'' \rrbracket = \llbracket f_{1} + \ldots + g_{p} + \ldots f_{k} \rrbracket = \llbracket g \rrbracket$ and $f = g'' = g$.
\end{itemize}
    % there is a sequence of SMC and enrichment equations such that when applied to $f$ it is turned into $g$.
    % As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \rrbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[]$ is defined by computing the pushout along the common interfaces.
    % We first note that any subterm of $g$ correspond to a convex down-closed subgraph of $\llbracket f \rrbracket$ and hence there is always a convex down-closed match of $\llbracket l \rrbracket$ within $\llbracket g \rrbracket$.
    % The mapping of external interfaces into $\llbracket l \rrbracket$ is mono and therefore the corresponding pushout complement exists.
    % We then need to show that for all $\llbracket l \rrbracket = \llbracket r \rrbracket$ in $\mathcal{S}$ the match satisfies the conditions in Proposition~\ref{prop:complement_existence} which will complete the first square of EDPOI and then check that the arrows in the second square satisfy the condition from Theorem~\ref{th:existence_of_pushouts}.

@@

        n \arrow[r] & n'_{1} \arrow[r] & \mathcal{C}_{1} & m'_{1} + k \arrow[l]     & i+k \arrow[l] \arrow[r] & i'+k \arrow[rd] &             \\

>>>>

    Then, faithfulness amounts to showing that if $\llbracket f \rrbracket \Rrightarrow_{\llbracket \mathcal{E} \rrbracket}^{*} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$ then $f = g$ modulo $\mathcal{E}$ in $\catname{S}^{+}(\Sigma)$.
% By the theorem~\ref{thm:completeness_simple} if $\llbracket f \rrbracket = \llbracket g \rrbracket$ via rules from $\mathcal{S}$ then $f = g$, so we will show that the same holds when $\llbracket f \rrbracket = \llbracket g \rrbracket$ via rules from $\mathcal{E}$.
Consider a rewrite rule $\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle = \langle \mathcal{L}, \mathcal{R} \rangle$ such that $\llbracket f \rrbracket \Rrightarrow_{\langle \mathcal{L}, \mathcal{R} \rangle} \llbracket g \rrbracket$ and suppose
\[
	\llbracket f \rrbracket = n \xrightarrow{} n' \xrightarrow{} \mathcal{F} \xleftarrow{} m' \xleftarrow{} m ~.
\]

\begin{figure*}
    \begin{subfigure}{\linewidth}
\begin{align*}
    (k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
    n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
    (i \xrightarrow{} i' \xrightarrow{} &\mathcal{L} \xleftarrow{} j' \xleftarrow{} j)
\end{align*}
\subcaption{\;}
\end{subfigure}
\vspace{1ex}
\begin{subfigure}{\linewidth}
    \begin{align*}
        (k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
        n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
        (i \xrightarrow{} i' \xrightarrow{} &\mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)
    \end{align*}
\subcaption{\;}
\end{subfigure}
\caption{Decomposition}
\label{fig:decomposition_1}
\end{figure*}

This implies the existence of a diagram as in~\ref{def:dpoi-e}.
We will proceed by induction on the level of the occurrence of $\mathcal{L}$ within $\mathcal{F}$.
\begin{itemize}
	\item Suppose the image of $\mathcal{L}$ in $\mathcal{F}$ is top-level.
		  Then, by lemma~\ref{lemma:decomposition}, there exists a decomposition of $\llbracket f \rrbracket$ in terms of MDA cospans as in Figure~\ref{fig:decomposition_1}.

		% \begin{align*}
		% 	(k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
		% 	n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
		% 	(i \xrightarrow{} i' \xrightarrow{} &\mathcal{L} \xleftarrow{} j' \xleftarrow{} j)
		% \end{align*}
		By fullness of $\llbracket - \rrbracket$ (from Theorem~\ref{thm:completeness_simple}) there exist $c_{1}$ and $c_{2}$ such that 
		\begin{align*}
		&\llbracket c_{1} \rrbracket = n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k\\
        &\llbracket c_{2} \rrbracket = j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m
        \end{align*}
		By functoriality, $\llbracket f \rrbracket = \llbracket c_{1};(id \otimes l);c_{2} \rrbracket$ and by faithfulness we have that $f = c_{1};(id \otimes l);c_{2}$ to which we can apply $\langle l, r \rangle$ and get $g' = c_{1};(id \otimes r);c_{2}$.
		We then have $\llbracket g' \rrbracket = \llbracket c_{1};(id \otimes r);c_{2} \rrbracket =  \llbracket g \rrbracket$ since the boundary complement is unique (Proposition ~\ref{prop:boundary_unique}) and $g' = g$ by faithfulness.

		\item If the image of $\mathcal{L}$ is not top-level, then either $\llbracket f \rrbracket$ is decomposable as in Figure~\ref{fig:decomposition_2}
		% \begin{align*}
		% 	(k \xrightarrow{id_{k}} k \xrightarrow{id_{k}} &k \xleftarrow{id_{k}} k \xleftarrow{id_{k}} k)\\
		% 	n \xrightarrow{} n_{1}' \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m_{1}' + i + k \xleftarrow{} i + k; \qquad\qquad\qquad &\otimes \qquad\qquad\qquad ; j + k \xrightarrow{} j + k + n'_{2} \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_{2} \xleftarrow{} m\\
		% 	(i \xrightarrow{} i' \xrightarrow{} &\mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)
		% \end{align*}
		such that $\mathcal{L}$ is a convex down-closed sub-e-hypergraph of $F'$.
		Since the image of $\mathcal{L}$ lies within $\mathcal{F}'$, there is an EDPO rewrite step that turns
		\[
			(i \xrightarrow{} i' \xrightarrow{} \mathcal{F'} \xleftarrow{} j' \xleftarrow{} j)	
		\]
		into
		\[
			(i \xrightarrow{} i'' \xrightarrow{} \mathcal{G'} \xleftarrow{} j'' \xleftarrow{} j)	
		\]
		via $\langle \mathcal{L}, \mathcal{R} \rangle$.
		By fullness and inductive hypothesis, there exist $f'$ and $g'$ such that 
		\begin{align*}
			&\llbracket f' \rrbracket = i \xrightarrow{} i' \xrightarrow{} \mathcal{F'} \xleftarrow{} j' \xleftarrow{} j\\
            &\llbracket g' \rrbracket = (i \xrightarrow{} i'' \xrightarrow{} \mathcal{G'} \xleftarrow{} j'' \xleftarrow{} j)
        \end{align*}
		and that $f' = g'$.
		By decomposition and fullness we have $\llbracket f \rrbracket = \llbracket c_{1};(id \otimes f');c_{2} \rrbracket$ and $f = c_{1};(id \otimes f');c_{2}$ by faithfulness.
		By applying $f' = g'$ we get $f = g'' = c_{1};(id \otimes g');c_{2}$.
		Finally, $\llbracket g'' \rrbracket = \llbracket c_{1};(id \otimes g');c_{2} \rrbracket = \llbracket g \rrbracket$ (by the uniqueness of the boundary complement) and $g'' = g = f$ by faithfulness.

		\item or, $\llbracket f \rrbracket$ is decomposable as
		\begin{align*}
			n_1 \xrightarrow{} n_1' \xrightarrow{} &\mathcal{F}_1 \xleftarrow{} m_1' \xleftarrow{} m_1\\
			&\;+\\
			&\vdotswithin{+}\\
			&\;+\\
			n_p \xrightarrow{} n_p' \xrightarrow{} &\mathcal{F}_p \xleftarrow{} m_p' \xleftarrow{} m_p\\
			&\;+\\
			&\vdotswithin{+}\\
			&\;+\\
			n_k \xrightarrow{} n_k' \xrightarrow{} &\mathcal{F}_k \xleftarrow{} m_k' \xleftarrow{} m_k
		\end{align*}
		such that the image of $\mathcal{L}$ lies within $\mathcal{F}_{p}$.
		By inductive hypothesis we have that $\llbracket f_{p} \rrbracket = n_p \xrightarrow{} n_{p}' \xrightarrow{} \mathcal{F}_{p} \xleftarrow{} m_{p}' \xleftarrow{} m_{p}$ rewrites into $\llbracket g_{p} \rrbracket = n_p \xrightarrow{} n_{p}'' \xrightarrow{} \mathcal{G}_{p} \xleftarrow{} m_{p}'' \xleftarrow{} m_{p}$ such that $f_{p} = g_{p}$.
		By fullness and the fact that $\llbracket - \rrbracket$ is $\catname{SLat}$-enriched, there exist $\llbracket f_1 + \ldots + f_{p} + \ldots + f_{k} \rrbracket = \llbracket f_{1} \rrbracket + \ldots + \llbracket f_{p} \rrbracket + \ldots + \llbracket f_{k} \rrbracket = \llbracket f \rrbracket$.
		By applying $f_{p} = g_{p}$ we get $f = g''$ and $\llbracket g'' \rrbracket = \llbracket f_{1} + \ldots + g_{p} + \ldots f_{k} \rrbracket = \llbracket g \rrbracket$ and $f = g'' = g$.
\end{itemize}
    % there is a sequence of SMC and enrichment equations such that when applied to $f$ it is turned into $g$.
    % As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \rrbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[]$ is defined by computing the pushout along the common interfaces.
    % We first note that any subterm of $g$ correspond to a convex down-closed subgraph of $\llbracket f \rrbracket$ and hence there is always a convex down-closed match of $\llbracket l \rrbracket$ within $\llbracket g \rrbracket$.
    % The mapping of external interfaces into $\llbracket l \rrbracket$ is mono and therefore the corresponding pushout complement exists.
    % We then need to show that for all $\llbracket l \rrbracket = \llbracket r \rrbracket$ in $\mathcal{S}$ the match satisfies the conditions in Proposition~\ref{prop:complement_existence} which will complete the first square of EDPOI and then check that the arrows in the second square satisfy the condition from Theorem~\ref{th:existence_of_pushouts}.

@@

        n \arrow[r] & n'_{1} \arrow[r] & \mathcal{C}_{1} & m'_{1} + i + k \arrow[l]     & i+k \arrow[l] \arrow[r] & i'+k \arrow[rd] &             \\
